---
title: "earth_eater"
author: "Tahmid Hossain Bhuiyan"
date: "`r Sys.Date()`"
output: ioslides_presentation
---


## Are we eating away our earth to survive?

**A deepdive on SDG-12**


- 8 billion+ people
- 3 meals a day
- 1 planet

## Act 1: 1B to 8B in ~ 200 years!

- Acceleration after 1950
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)

# Import global population dataset from Our World in Data
population <- read_csv('https://ourworldindata.org/grapher/population.csv?v=1&csvType=full&useColumnShortNames=true')

# Clean and prepare dataset:
# Rename columns for clarity
# Filter only "World" data to analyze global trends
# Restrict timeline to post-1800 to highlight modern acceleration

pop<- population %>% rename(country=entity,
                            population=population_historical) %>% 
  filter(country=="World",
         year>1800)

# Create population growth visualization using ggplot.

p1 <- ggplot(pop,
       aes(x= year,
           y=population)) +
  geom_line(linewidth=0.5,
            color="red") +
  labs(title = "Humanity’s Growth Has Been Exponential",
    subtitle = "Global population growth, 1800–2023",
    y = "Population",
    x = "Year",
    caption = "Source: Our World in Data") +
  
  # Format y-axis into readable billions/millions
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  
  # Highlight 1950 as structural break (post-WWII growth era)
  geom_vline(xintercept = 1950,
             linetype="dashed",
             color="orange",
             linewidth=0.7,
             alpha=0.9)+
  annotate("text",
           x=1950,
           y=2.9e9,
           label="1950:Population Boom era")+
  theme_minimal()

# Convert to interactive plot for enhanced engagement
ggplotly(p1)
```


## Act 2: We Don’t Just Multiply. We Consume More.
- More People. More Calories. More Pressure

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)

#########################
# Re-import population dataset to combine with calorie data

population <- read_csv('https://ourworldindata.org/grapher/population.csv?v=1&csvType=full&useColumnShortNames=true')


# Data Cleaning
pop<- population %>% rename(country=entity,
                            population=population_historical) %>% 
  filter(country=="World",
         year>1800)

########################
# Import per-capita daily calorie supply dataset
# This captures consumption intensity 

calorie <- read_csv('https://ourworldindata.org/grapher/daily-per-capita-caloric-supply.csv?v=1&csvType=full&useColumnShortNames=true')

cal <- calorie %>% rename(country=entity) %>% 
  filter(country=="World",
         year>1800)

# Merge datasets by year
# inner_join ensures only overlapping years are used 

food_demand<- cal %>% inner_join(pop,
                                 by="year")

# Total global daily calories = per capita calories × population
# Divided by 1e12 to express in trillions

food_demand2 <- food_demand %>% 
  mutate(total_calories_trillions=daily_calories*population/1e12) %>% 
  select(-c(country.y, code.y)) # Remove redundant columns after join

# Visualize total global food demand

ggplot(food_demand2,
       aes(x=year,
           y=total_calories_trillions)) +
  geom_line(color='orange',
            alpha=0.6) +
  labs(title = "Global Food Demand Has Exploded",
    subtitle = "Total daily calories required worldwide",
    x = "Year",
    y = "Trillions of calories per day",
    caption = "Source: Our World in Data | Population × Daily calorie supply"
  ) +
  # Focus x-axis on modern decades where growth accelerates
  scale_x_continuous(breaks=c(1960,1970,1980,1990,2000,2010,2020)) +
  theme_minimal()
```

## Act 3: The Land Responds

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
#install.packages('forcats') # Installing forcats package for fct_reorder
library(forcats)

# Import long-term agricultural land data

agriland <- read_csv('https://ourworldindata.org/grapher/total-agricultural-area-over-the-long-term.csv?v=1&csvType=full&useColumnShortNames=true')

# Select key regions + global aggregate
countries <- c("World", "Asia", "Europe", "Africa", "North America")

agri <- agriland %>% rename(country=entity,
                            agriculture_area=agriculture_c) %>%
  filter(country %in% countries,
         year>1800) %>% 
# Reorder factor levels for clearer hierarchy
  mutate(country=fct_reorder(country, desc(agriculture_area)))

# Area chart with ggplot
ggplot(agri,
       aes(x=year,
           y=agriculture_area,
           fill=country)) %>% +
  geom_area(alpha=0.8) +
  scale_y_continuous(labels=scales::label_number(scale_cut=scales::cut_short_scale()))+
  scale_x_continuous(expand=c(0,0), # So that the graph starts from 1810 without gap
                     breaks=c(1810,1850,1900,1950,1990,2020))+
  labs(
    title = "To Feed Ourselves, We Reshaped the Planet",
    subtitle = "Agricultural land over time",
    x = "Year",
    y = "Agricultural land (hectares)",
    caption = "Source: Our World in Data")+
  theme_minimal()
```

## Act 3: What We Gained in Farms, We Lost in Forests

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
#install.packages('forcats')
library(forcats)
library(plotly)

forestland <- read_csv('https://ourworldindata.org/grapher/forest-area-as-share-of-land-area.csv?v=1&csvType=full&useColumnShortNames=true')

forest <- forestland %>% 
  rename(country=entity) %>% 
  filter(country=="World",
         year>1800) %>% 
  mutate(country=fct_reorder(country, desc(forest_share)))

# Area + line overlay emphasizes long-term downward trend
f1 <- ggplot(forest,
       aes(x=year,
           y=forest_share))+
  geom_area(fill = "green", alpha = 0.3) +
  geom_line(color = "#1B5E20", linewidth = 1.2)+
  labs(title = "Slow but steady forest share decline",
       subtitle = "Global forest area as share of total land",
       x = "Year",
       y = "Forest share(%)",
       caption = "Source: Our World in Data") +
  theme_minimal()

ggplotly(f1)
```

## Act 4: The Atmosphere Answers Back

- Average temperature started rising rapidly.
- And this has a name called "Global Warming"!

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)

temperature <- read_csv('https://ourworldindata.org/grapher/temperature-anomaly.csv?v=1&csvType=full&useColumnShortNames=true')

# Reshape dataset for multi-line comparison (average, upper, lower bounds)
# Used pivot longer
temp <- temperature %>% 
  rename(country=entity,
         upper_temp=near_surface_temperature_anomaly_upper,
         average_temp=near_surface_temperature_anomaly,
         lower_temp=near_surface_temperature_anomaly_lower) %>% 
  filter(country=="World") %>% 
  pivot_longer(cols=c(average_temp,upper_temp,lower_temp),
               names_to="temp_level",
               values_to = "temp_degree")

ggplot(temp,
       aes(x=year,
           y=temp_degree,
           color=temp_level))+
  geom_line()+
# Mark mid-20th century acceleration period
  geom_vline(xintercept=1964,
             linetype="dashed",
             color="orange")+
  annotate("text",
           x=1964,
           y=-0.1,
           label="1964:Rapid warming era begins",
           size=3.4)+
  scale_color_manual(values = c(
    "average_temp" = "red",
    "upper_temp"   = "gray50",
    "lower_temp"   = "gray50")) +
  labs(
    title = "The Bill Comes Due",
    subtitle = "Global near-surface temperature anomaly, 1850–2023",
    x = "Year",
    y = "Temperature anomaly (°C)",
    caption = "Source: Our World in Data. Temperature anomaly relative to pre-industrial baseline.") +
  theme_minimal()
```

## Act 5: So.... what's the answer?

The answer is YES. 

We are eating away our earth to survive.

But is it because we are too many people?

- NO

Then it must be because we humans eat too much?

- Well, NOPE!

Then?

- It is because we don't know "HOW" to eat sustainably.


